Gravitaitonal Wave Event Progenitor Synthesis

### 2022-05-26 src/tex/ms.tex

Once all the parameters including a particular draw of all the random variables
represented by $\bm{X}$ are known, the population synthesis code is a
deterministic function that transforms the properties of the progenitors into
the GW parameters $\bm{\theta}$:
\begin{equation}
    \bm{\theta} = \bm{f}\left( \bm{\Theta} \right).
\end{equation}
The mapping $\bm{f}$ from $\bm{\Theta}$ to $\bm{\theta}$ can be many-to-one
due to degeneracies in the different physical processes and initial parameters
of the GW event progenitors.
In order to draw inferences about $\bm{\Theta}$ from GW data, we
must be able to evaluate the likelihood of that data at fixed progenitor
parameters, namely
\begin{equation}
    p\left( d \mid \bm{\Theta} \right) = p\left( d \mid \bm{\theta} = \bm{f}\left( \bm{\Theta} \right) \right).
\end{equation}
The equality holds because the likelihood depends only on the gravitational waveform generated by
parameters $\bm{\theta}$ to which the progenitor parameters map onto.  In principle
this likelihood could be computed at arbitrary values of $\bm{\Theta}$ using the
same machinery that is used to estimate source parameters in GW
catalogs \citep{Veitch2015,Ashton2019,Romero-Shaw2020,GWTC-3} to develop samples
of progenitor parameters, hyperparameters, and random variables in
$\bm{\Theta}$ from a posterior density.

Since the GW likelihood function
ultimately depends only on the GW parameters $\theta$, and GWTC3 already has samples over these parameters from a posterior density
\begin{equation}
    p\left( \bm{\theta} \mid d \right) \propto p\left( d \mid \bm{\theta} \right) \pi_\mathrm{GW} \left( \bm{\theta} \right),
\end{equation}
where $\pi_\mathrm{GW}$ is the prior density used for sampling,
we can save the computation cost associated with evaluating the GW likelihood by rewriting the posterior
density of $\Theta$ in terms of the posterior density of $\theta$, namely

\begin{align}
    p(\Theta | d) = \frac{p(\theta(\Theta)| d) \pi(\Theta)}{\pi_\mathrm{GW}(\theta(\Theta))},
\end{align}
which we apply a kernel density estimator to the posterior samples released in GWTC3 to estimate $p(\theta(\Theta)|d)$.

Including hyperparameters and random variables %in the sampling
drastically increases the dimensionality of the problem,
which makes the sampling process more computationally expensive to converge.
To speed up the convergence, we first solve a root-finding problem to find points that are close to the bulk of the posterior density of a GW event,
then we use those points to initialize a set of chains in the MCMC process.

@@

Retaining explicit control over random variables allows us to marginalize over their contribution,
and thus allows us to focus on progenitor parameters and hyperparameters.
However, in practice most population synthesis codes (including \texttt{COSMIC} which we use in this study)
define their random variables implicitly such that the random variables are drawn within the program instead of
being passed as arguments. This means that the process of evolving the binary is not fully deterministic.
Thus, even if the root-finding algorithm performs perfectly, forward modelling a set of roots does not
guarantee that the simulated population reproduces exactly the set of posterior samples due to
randomness in the evolution of each binary. To assure that the recovered progenitor and hyperparameters
robustly correspond to the posterior in the GW-observable space, we forward model (or "reproject") the recovered
evolutionary parameters to the GW-observable space to check whether the reprojected posterior agrees
with the posterior given by the LVK collaboration.

>>>>

Once all the parameters including a particular draw of all the random variables
represented by $\bm{X}$ are known, the population synthesis code is a
deterministic function that transforms the properties of the progenitors into
the GW parameters $\bm{\theta}$:
\begin{equation}
    \bm{\theta} = \bm{f}\left( \bm{\Theta} \right).
\end{equation}
The mapping $\bm{f}$ from $\bm{\Theta}$ to $\bm{\theta}$ can be many-to-one
due to degeneracies in the different physical processes and initial parameters
of the GW event progenitors.
In order to draw inferences about $\bm{\Theta}$ from GW data, we
must be able to evaluate the likelihood of that data at fixed progenitor
parameters, namely
\begin{equation}
    p\left( d \mid \bm{\Theta} \right) = p\left( d \mid \bm{\theta} = \bm{f}\left( \bm{\Theta} \right) \right).
\end{equation}
The equality holds because the likelihood depends only on the gravitational
waveform generated by parameters $\bm{\theta}$ onto which the progenitor
parameters map.  In principle this likelihood could be computed at arbitrary
values of $\bm{\Theta}$ using the same machinery that is used to estimate source
parameters in GW catalogs \citep{Veitch2015,Ashton2019,Romero-Shaw2020,GWTC-3}
to develop samples of progenitor parameters, hyperparameters, and random
variables in $\bm{\Theta}$ from a posterior density.

Since the GW likelihood function
ultimately depends only on the GW parameters $\theta$, and GWTC3 already has samples over these parameters from a posterior density
\begin{equation}
    p\left( \bm{\theta} \mid d \right) \propto p\left( d \mid \bm{\theta} \right) \pi_\mathrm{GW} \left( \bm{\theta} \right),
\end{equation}
where $\pi_\mathrm{GW}$ is the prior density used for sampling, we can save the
computational cost associated with evaluating the GW likelihood by rewriting the
posterior density of $\bm{\Theta}$ in terms of the posterior density of
$\bm{\theta}$, namely

\begin{align}
    p(\bm{\Theta} | d) = \frac{p(\bm{\theta}(\bm{\Theta})| d) \pi(\bm{\Theta})}{\pi_\mathrm{GW}(\bm{\theta}(\bm{\Theta}))}.
\end{align}
We apply a kernel density estimator with a Gaussian kernel to the posterior
samples released in GWTC3 to estimate $p(\bm{\theta}(\bm{\Theta})|d)$.

Including hyperparameters and random variables %in the sampling
drastically increases the dimensionality of the problem, which can make the
sampling process much more computationally expensive to converge. To speed up
the convergence, we first solve a root-finding problem to find points
$\bm{\Theta}$ that will generate GW parameters $\bm{\theta}$ close to the bulk
of the posterior density of a GW event.  We then use those points to initialize
a set of chains in the MCMC process.

@@

Retaining explicit control over random variables allows us to marginalize over
their contribution, and thus allows us to focus on progenitor parameters and
hyperparameters. However, in practice most population synthesis codes (including
\texttt{COSMIC} which we use in this study) define their random variables
implicitly such that the random variables are drawn within the program instead
of being passed as arguments. This means that the process of evolving the binary
is not fully deterministic. Thus, even if the root-finding algorithm performs
perfectly, forward modelling a set of roots does not guarantee that the
simulated population reproduces exactly the set of posterior samples due to
randomness in the evolution of each binary. To assure that the recovered
progenitor and hyperparameters robustly correspond to the posterior in the
GW-observable space, we push forward (or ``reproject'') the recovered
evolutionary parameters to the GW-observable space to check whether the
reprojected posterior agrees with the posterior given by the LVK collaboration.
